\documentclass[11pt,a4paper]{article}

%%% Packages %%%
\usepackage{amsmath,amsthm,amssymb,amsfonts}
\usepackage{booktabs}
\usepackage{mathrsfs}
\usepackage{tikz}
\usetikzlibrary{matrix,arrows,decorations.pathmorphing,positioning,calc}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{xcolor}

%%% Page geometry %%%
\usepackage[margin=1in]{geometry}

%%% Theorem environments %%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{question}[theorem]{Question}
\newtheorem{conjecture}[theorem]{Conjecture}

%%% Macros %%%
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\gl}{\mathfrak{gl}}
\newcommand{\Sym}{\mathrm{Sym}}
\newcommand{\Rep}{\mathrm{Rep}}
\newcommand{\Crys}{\mathrm{Crys}}
\newcommand{\Vect}{\mathrm{Vect}}
\newcommand{\Set}{\mathrm{Set}}

%%% Title %%%
\title{Littlewood--Richardson Coefficients:\\
  From Puzzles to the Integrability Hierarchy}
\author{[To be determined]}
\date{April 2026}

\begin{document}

\maketitle

\begin{abstract}
% TODO: Write abstract.
%
% The abstract should cover:
% - LR coefficients as the central object (multiplicities in tensor products,
%   Schubert calculus, symmetric function structure constants)
% - The puzzle/honeycomb combinatorial perspective (Knutson--Tao--Woodward)
% - The integrable lattice model perspective (Zinn-Justin, Wheeler--Zinn-Justin)
% - The hierarchy: Yang--Baxter (braided) -> coboundary (cactus) -> crystal (symmetric)
%   as successive forgetting of structure
% - The categorical phase diagram organizing these levels
% - How every known combinatorial rule for LR coefficients fits into
%   this integrability hierarchy
%
Littlewood--Richardson coefficients $c_{\lambda\mu}^{\nu}$ govern the
multiplication of Schur functions, the decomposition of tensor products
of $\GL_n$ representations, and the intersection theory of Grassmannians.
We survey the landscape of combinatorial rules for these coefficients---from
the classical Littlewood--Richardson rule through Knutson--Tao--Woodward
puzzles to integrable lattice models---and organize them into a hierarchy
controlled by the Yang--Baxter equation and its degenerations.
At the top sits the fully solvable regime (braided monoidal, spectral parameter,
$R$-matrices); in the middle, the quasi-solvable or coboundary regime
(cactus group, crystals with commutor); at the bottom, the symmetric
monoidal regime (bare crystals, no spectral data). We argue that every
known combinatorial rule for $c_{\lambda\mu}^{\nu}$ and its
$K$-theoretic and equivariant generalizations finds a natural home in
this categorical phase diagram, and we identify the open problems that
remain.
% TODO: Revise and polish once all sections are complete.
\end{abstract}

\tableofcontents

\bigskip

%%% Section files %%%
\input{section1-introduction}
\input{section2-classical}
\input{section3-lattice}
\input{section4-groupoid}
\input{section5-hierarchy}
\input{section6-shuffle}
\input{section7-open}

%%% Acknowledgements %%%
\section*{Acknowledgements}
% TODO: Add acknowledgements.

%%% Bibliography %%%
\begin{thebibliography}{99}

\bibitem{clio-midpoint-2026}
[Author omitted for review],
\emph{The cactus midpoint theorem: braid obstruction and word-independence
in the 0-Hecke algebra},
manuscript, April 2026.

\bibitem{clio-h-invariant-2026}
[Author omitted for review],
\emph{The H-invariant rank theorem for the Hecke staircase product},
in preparation, 2026.

\bibitem{clio-eigenvalue-positivity-2026}
[Author omitted for review],
\emph{Eigenvalue positivity of Hecke products via Kazhdan--Lusztig W-graphs},
in preparation, 2026.

\bibitem{clio-first-pair-2026}
[Author omitted for review],
\emph{The first $q$-shifted pair: $\det M_R^{(3)}[D_n] = (\det M_R^{(2)}[D_n])^2$},
manuscript, April 2026.

\bibitem{clio-second-pair-2026}
[Author omitted for review],
\emph{The second $q$-shifted pair: rigorous closure at $n = 6$},
manuscript, April 2026.

\bibitem{clio-uniform-locality-2026}
[Author omitted for review],
\emph{Uniform locality for the staircase determinants $\det M_R^{(k)}[D_n]$},
manuscript, April 2026.

\bibitem{sandvik-2026}
G.~Sandvik,
\emph{The monodromic Hecke category and parity sheaves on $U \backslash G / U$},
arXiv:2604.15582 (2026).

\bibitem{woo-mo-2024}
A.~Woo,
\emph{Comment on ``What do Kazhdan--Lusztig polynomials measure?''},
MathOverflow Q477404 (2024),
\url{https://mathoverflow.net/q/477404}.

\bibitem{alqady-stroinski-2025}
R.~Alqady and M.~Stroinski,
\emph{The Temperley--Lieb category as a coboundary category},
J. London Math.\ Soc., to appear (2025).

\bibitem{azenhas-feller-torres-2022}
O.~Azenhas, P.~L.~Feller, and J.~Torres,
\emph{Symplectic cacti, virtualization, and Berenstein--Zelevinsky data},
J. Comb.\ Algebra, to appear (2026).

\bibitem{baxter-1982}
R.~J.~Baxter,
\emph{Exactly Solved Models in Statistical Mechanics},
Academic Press, London, 1982.

\bibitem{borodin-wheeler-2018}
A.~Borodin and M.~Wheeler,
\emph{Coloured stochastic vertex models and their spectral theory},
arXiv:1808.01866 (2018).

\bibitem{brubaker-bump-buciumas-gustafsson-2021}
B.~Brubaker, D.~Bump, V.~Buciumas, and H.~P.~A.~Gustafsson,
\emph{Colored five-vertex models and Demazure atoms},
J. Combin.\ Theory Ser.~A \textbf{178} (2021), 105354.

\bibitem{brubaker-bump-hardt-spink}
B.~Brubaker, D.~Bump, A.~Hardt, and H.~Spink,
\emph{Kazhdan--Lusztig theory and the Yang--Baxter equation},
arXiv preprint (2024).

\bibitem{buciumas-scrimshaw}
V.~Buciumas and T.~Scrimshaw,
\emph{Quasi-solvable lattice models for $\mathrm{Sp}_{2n}$ and $\mathrm{SO}_{2n+1}$ Demazure atoms and characters},
arXiv:2201.01026 (2022).

\bibitem{bump-naprienko-2025}
D.~Bump and S.~Naprienko,
\emph{The Yang--Baxter equation and Schur functions},
arXiv preprint (2025).

\bibitem{devyatov-2015}
R.~Devyatov,
\emph{On subgroups of the pure cactus group},
arXiv:1507.04699 (2015).

\bibitem{fan-guo-xiong}
N.~Fan, P.~Guo, and R.~Xiong,
\emph{Bumpless pipe dreams meet puzzles},
arXiv preprint (2024).

\bibitem{felder-1995}
G.~Felder,
\emph{Conformal field theory and integrable systems associated to elliptic curves},
Proc.\ ICM, Z\"urich (1994), Birkh\"auser, 1995, pp.~1247--1255.

\bibitem{fulton-1997}
W.~Fulton,
\emph{Young Tableaux: With Applications to Representation Theory and Geometry},
London Math.\ Soc.\ Student Texts \textbf{35}, Cambridge Univ.\ Press, 1997.

\bibitem{gaetz-gao-2024}
C.~Gaetz and Y.~Gao,
\emph{Integrable triangle-accessible lattice models},
arXiv preprint (2024).

\bibitem{galashin-2020}
P.~Galashin,
\emph{Symmetries of stochastic colored vertex models},
Ann.\ Probab.\ \textbf{49} (2021), no.~5, 2175--2219.

\bibitem{llt-1997}
A.~Lascoux, B.~Leclerc, and J.-Y.~Thibon,
\emph{Hecke algebras at roots of unity and crystal bases of quantum affine algebras},
Comm.\ Math.\ Phys.\ \textbf{181} (1996), no.~1, 205--263.

\bibitem{garbali-gunna-2024}
A.~Garbali and J.~Gunna,
\emph{Shuffle algebras and lattice paths},
arXiv preprint (2024).

\bibitem{garbali-negut-2025}
A.~Garbali and A.~Negut,
\emph{Shuffle algebras and Yangians},
arXiv preprint (2025).

\bibitem{gorbounov-korff-mihalcea-2025}
V.~Gorbounov, C.~Korff, and L.~Mihalcea,
\emph{Quantum $K$-theory and the Yang--Baxter equation},
arXiv preprint (2025).

\bibitem{gunna-wheeler-zj-2025}
J.~Gunna, M.~Wheeler, and P.~Zinn-Justin,
\emph{Spin Hall--Littlewood polynomials and vertex models},
arXiv preprint (2025).

\bibitem{henriques-kamnitzer-2006}
A.~Henriques and J.~Kamnitzer,
\emph{Crystals and coboundary categories},
Duke Math.~J.\ \textbf{132} (2006), no.~2, 191--216.

\bibitem{jimbo-1986}
M.~Jimbo,
\emph{A $q$-analogue of $U(\mathfrak{gl}(N+1))$, Hecke algebra,
and the Yang--Baxter equation},
Lett.\ Math.\ Phys.\ \textbf{11} (1986), no.~3, 247--252.

\bibitem{kalmykov}
A.~Kalmykov,
\emph{Shuffle algebra, degenerate affine Hecke algebra, and Yangian},
preprint.

\bibitem{kamnitzer-tingley-2007}
J.~Kamnitzer and P.~Tingley,
\emph{The crystal commutor and Drinfeld's unitarized $R$-matrix},
J.~Algebraic Combin.\ \textbf{29} (2009), no.~3, 315--335.

\bibitem{kashiwara-1990}
M.~Kashiwara,
\emph{Crystallizing the $q$-analogue of universal enveloping algebras},
Comm.\ Math.\ Phys.\ \textbf{133} (1990), no.~2, 249--260.

\bibitem{knutson-tao-1999}
A.~Knutson and T.~Tao,
\emph{The honeycomb model of $\mathrm{GL}_n(\mathbb{C})$ tensor products~I:
Proof of the saturation conjecture},
J.~Amer.\ Math.\ Soc.\ \textbf{12} (1999), no.~4, 1055--1090.

\bibitem{knutson-tao-woodward-2004}
A.~Knutson, T.~Tao, and C.~Woodward,
\emph{The honeycomb and saturation conjectures},
J.~Amer.\ Math.\ Soc.\ \textbf{17} (2004), no.~1, 19--48.

\bibitem{macdonald-1995}
I.~G.~Macdonald,
\emph{Symmetric Functions and Hall Polynomials},
2nd ed., Oxford Math.\ Monographs, Oxford Univ.\ Press, 1995.

\bibitem{morales-panova-petrov-yeliussizov}
A.~H.~Morales, G.~Panova, L.~Petrov, and D.~Yeliussizov,
\emph{Grothendieck shenanigans},
arXiv preprint (2024).

\bibitem{norton-1979}
P.~N.~Norton,
\emph{$0$-Hecke algebras},
J.~Austral.\ Math.\ Soc.\ Ser.~A \textbf{27} (1979), no.~3, 337--357.

\bibitem{paten-woodruff-2025}
J.~Paten and E.~Woodruff,
\emph{Key positivity and $0$-Hecke characters},
preprint, arXiv:2503.xxxxx (2025).

\bibitem{purbhoo-2007}
K.~Purbhoo,
\emph{Puzzles, tableaux, and mosaics},
J.~Algebraic Combin.\ \textbf{28} (2008), no.~4, 461--480.

\bibitem{savage-2020}
A.~Savage,
\emph{Crystals, coboundary categories, and the commutor},
lecture notes (2020).

\bibitem{shibukawa-ueno}
Y.~Shibukawa and K.~Ueno,
\emph{Completely $\mathbb{Z}$-symmetric R-matrices},
Lett.\ Math.\ Phys.\ \textbf{25} (1992), no.~3, 239--248.

\bibitem{stanley-1999}
R.~P.~Stanley,
\emph{Enumerative Combinatorics}, vol.~2,
Cambridge Studies in Adv.\ Math.\ \textbf{62}, Cambridge Univ.\ Press, 1999.

\bibitem{tikhonov-2026}
K.~Tikhonov,
\emph{Random permutations from $q$-Demazure products},
arXiv:2604.06532 (2026).

\bibitem{wheeler-zinn-justin-2016}
M.~Wheeler and P.~Zinn-Justin,
\emph{Littlewood--Richardson coefficients for Grothendieck polynomials
from integrability},
J.~Reine Angew.\ Math.\ \textbf{757} (2019), 159--195.

\bibitem{yang-2025}
J.~Yang,
\emph{Closed colored lattice models and Demazure crystals},
arXiv:2512.05479 (2025).

\end{thebibliography}

\end{document}
